Naturalness and Long-Lived Beyond the Standard Model Particles

### 2017-08-01 tex/draft.tex
Major edit of Viability

    Experimentally, the $\Wpm$, $\Z$, quarks, and charged leptons are massive\fxnote{Gloss over neutrino mass?}. Explicit mass terms for the fermions or gauge bosons would violate the chiral and gauge symmetries of the SM Lagrangian, however, so $\SUtwoLUoneY$ must be broken. Specifically, the nonzero vacuum expectation value (VEV) of the Higgs field spontaneously breaks $\SUtwoLUoneY$ to $\UoneQ$. In the process, three of the four degrees of freedom associated with the complex Higgs doublet become the longitudinal polarizations of $\Wpm$ and $\Z$, giving them mass. The remaining degree of freedom is the physical Higgs boson. Expanding the Higgs field about its VEV produces Yukawa interactions such as those in Eq.~\ref{yukawa_terms}. These terms are interpreted as giving mass to the quarks and charged leptons mass, where the mass of each particle is given by the product of the Higgs VEV, $v$, and the coupling constant, $Y$.

    \noindent \begin{equation}
        \mathcal{L} = -\frac{Yv}{\sqrt{2}}\bar{\psi}\psi 
        \label{yukawa_terms}
    \end{equation}

@@

    The naturalness criterion can be formulated in many ways, a few of which are reproduced here: (1) All dimensionless parameters should be of order one, (2) An effective theory should not be sensitive to small variations in the parameters of the higher-energy theory, and (3) any dimensionless parameter much smaller than unity must be protected by a custodial symmetry. Physicists implicitly, and perhaps unwittingly, apply naturalness in the form of (1) any time they use dimensional analysis to estimate a quantity, while (2) merely restates the idea that a predictive effective theory requires low-energy phenomena to be decoupled from the details of the high-energy theory\fxnote{This is my weakest point because I don't understand it as well as I'd like to. It's also wordy and maybe repetative}. This discussion focuses on statement (3), which is in some ways a more nuanced version of (1) that explains cases where dimensional analysis fails.

@@

    \noindent \begin{figure}[htbp] \begin{center}
        \feynmandiagram [inline=(d.base), layered layout, medium, horizontal=b to c] {
        a -- [scalar] b,
        b -- [fermion, half left] c,
        c -- [fermion, half left] b,
        c -- [scalar] d,
    };
    \quad + \quad
    \feynmandiagram [baseline={(current bounding box.center)}, layered layout, medium, horizontal=b to c] {
        a -- [scalar] b,
        b -- [charged scalar, half left] c,
        c -- [charged scalar, half left] b,
        c -- [scalar] d,
    };
    \quad = \quad 0 \quad
    \caption{Caption!\fxnote{cite something}}
    \label{exact_susy_correction}
    \end{center} \end{figure}

@@

    \noindent \begin{equation}
        \delta \mHiggs^2 = \frac{3Y^2}{8\pi^2}\left(m_{\sfermion}^2 - m_{\f}^2 \right)\ln(\Lambda^2/m_{\sfermion}^2).
    \label{higgs_susy_correction}
    \end{equation}

@@

    \subsection{Disappearing Tracks}
        \fxnote{cite CONF-2017-017}
        Certain models with nearly mass-degenerate LSP and NLSP have the potential to present a remarkable signal: charged particles that seem to disappear in the tracker. Specifically, models with winos as the lightest gauginos\fxnote{have these words been defined?}, such as many AMSB models\fxnote{why amsb? what is amsb?}, will have a chargino NLSP with an observable lifetime (\SI{0.2}{\nano\s} is a typical estimate\fxnote{why? how typical?}) that eventaully decays to the neutralino LSP and, commonly, a pion\fxnote{why?}. The nearly-degenerate masses have two important phenomenological roles: (1) increase the chargino lifetime, and (2) ensure low-momentum decay products. The low-momentum pion is then unlikely to be reconstructed, which results in a disappearing track.

@@

    Interpreting current limits in the context of naturalness requires some set of criteria that determine if a model is natural. No well-defined, universally-agreed-upon criteria exist, but certain themes recur in many discussions of natural model building\fxnote{better be able to confidently cite several sources}. The following discussion lays out a simplistic set of criteria for natural SUSY models and attempts to assess the status of naturalness in light of current experimental limits\fxnote{Did I just steal Chris's langauge?}.

@@

    Natural SUSY scenarios, then, are most likely to be constrained by limits on the stop, gluino, and Higgsino masses. The Higgsino, however, does not carry color charge and is far less likely\fxnote{be able to quantify} to be produced at the LHC, leaving stop and gluino mass limits as the most important contraints on natural SUSY models. As shown in \fxnote{reference correction terms}, the size of stop and gluino corrections also depends on the messenger scale of SUSY breaking.\fxnote{Make sure I understand this. Messenger scale != breaking scale?}\fxnote{Either explain more thoroughly or drop completely. What are the consequences of a lower messenger scale?}

    Two qualitative decisions must be made when analyzing the state of naturalness: (1) How best to define the degree of fine tuning in a particular model and (2) what degree of fine tuning is acceptable? A common, but not universal, definition for (1) is \fxnote{insert Delta eq.}, where $\alpha_i$ is a particular parameter and $\Delta_i$ is the degree of fine tuning from variations in that parameter. The overall degree of fine tuning for the model can then be assigned as $\max{\left(\Delta_i\right)}$, the product of $\Delta_i$, or all $\Delta_i$ added in quadrature. Determining the acceptable degree of fine tuning is even more open ended. Before null results and serious constraints started rolling in, it was common\fxnote{cite something} to refer to the $\Delta \lesssim 10$ regions of parameter space as natural. As models are constrained, physicists seem willing to accept greater degrees of fine tuning. Instead of asking, ``what degree of fine tuning should be considered natural?", Feng\fxnote{cite} proposes physicists ask themselves, ``If SUSY were discovered with such-and-such degree of fine tuning, would I consider the naturalness problem solved and move on to different research questions?" Anecdotally, Feng reports that acceptable levels of fine tuning tend to increase when the question is posed in this way.\fxnote{break into two paragraphs?}

    Fig.~\ref{atlas_susy_searches} summarizes limits on SUSY models from ATLAS searches. CMS limits are generally similar\fxnote{make sure this is true}, but a recent plot was not available.

@@

    Indeed, limits on mStop and mGluino from conventional SUSY searches wipe out much of the natural parameter space from prompt decays\fxnote{quantify? That would require getting into how one quantifies fine tuning} for many models with decoupled first and second geration fermions, baryonic RPV, or compressed spectra that minimize missing energy while the natural MSSM is completely excluded.\fxnote{Good spot for sweet Cornering plots}\fxnote{Use ATLAS SUSY summary plot too. CMS's isn't updated}

@@

    Despite strong contraints, SUSY remains one of the least-tuned options for BSM naturalness\fxnote{Make sure I can explain this and make sure it isn't plagarized from Cornering}.\fxnote{Good place for Feng ``how much tuning is acceptable? discussion}

@@

\clearpage
\pagebreak
\singlespacing
\bibliography{candidacy}{}
\bibliographystyle{auto}

>>>>

    Experimentally, the $\Wpm$, $\Z$, quarks, and charged leptons are massive\fxnote{Gloss over neutrino mass?}. Explicit mass terms for the fermions or gauge bosons would violate the chiral and gauge symmetries of the SM Lagrangian, however, so $\SUtwoLUoneY$ must be broken. Specifically, the nonzero vacuum expectation value (VEV) of the Higgs field spontaneously breaks $\SUtwoLUoneY$ to $\UoneQ$. In the process, three of the four degrees of freedom associated with the complex Higgs doublet become the longitudinal polarizations of $\Wpm$ and $\Z$, giving them mass. The remaining degree of freedom is the physical Higgs boson. Expanding the Higgs field about its VEV produces Yukawa interactions such as those in Eq.~\ref{yukawa_terms}. These terms are interpreted as giving mass to the quarks and charged leptons mass, where the mass of each particle is given by the product of the Higgs VEV, $v$, and the coupling constant, $y$.

    \noindent \begin{equation}
        \mathcal{L} = -\frac{yv}{\sqrt{2}}\bar{\psi}\psi 
        \label{yukawa_terms}
    \end{equation}

@@

    The naturalness criterion can be formulated in many ways, a few of which are reproduced here: (1) All dimensionless parameters should be of order one, (2) An effective theory should not be sensitive to small variations in the parameters of the higher-energy theory, and (3) any dimensionless parameter much smaller than unity must be protected by a custodial symmetry. Physicists implicitly, and perhaps unwittingly, apply naturalness in the form of (1) any time they use dimensional analysis to estimate a quantity, while (2) merely restates\fxnote{verb-subject agreement} the idea that a predictive effective theory requires low-energy phenomena to be decoupled from the details of the high-energy theory\fxnote{This is my weakest point because I don't understand it as well as I'd like to. It's also wordy and maybe repetative}. This discussion focuses on statement (3), which is in some ways a more nuanced version of (1) that explains cases where dimensional analysis fails.

@@

    \fxnote{Fix diagram} 
    \noindent \begin{figure}[htbp] \begin{center}
        \feynmandiagram [inline=(d.base), layered layout, medium, horizontal=b to c] {
        a -- [scalar] b,
        b -- [fermion, half left] c,
        c -- [fermion, half left] b,
        c -- [scalar] d,
    };
    \quad + \quad
        \begin{tikzpicture}
            \begin{feynman}
                \diagram [baseline={(current bounding box.center)}, horizontal=a to c] {
                    a -- [scalar] b,
                    b -- [scalar] c,
            };
                \draw (b) arc [charged scalar, start angle=270, end angle=90, radius=0.7cm];
            \end{feynman}
        \end{tikzpicture}
    \quad = \quad 0 \quad
    \caption{Broken!\fxnote{cite something}}
    \label{exact_susy_correction}
    \end{center} \end{figure}

@@

    \noindent \begin{equation}
        \delta \mHiggs^2 = \frac{3y_t^2}{8\pi^2}\left(m_{\sfermion}^2 - m_{\f}^2 \right)\ln(\Lambda^2/m_{\sfermion}^2).
    \label{higgs_susy_correction}
    \end{equation}

@@

    \subsection{Disappearing Tracks}
        \fxnote{cite CONF-2017-017}
        Certain models with nearly mass-degenerate LSP and NLSP have the potential to present a remarkable signal: charged particles that seem to disappear in the tracker. Specifically, models with winos as the lightest gauginos\fxnote{have these words been defined?}, such as many AMSB models\fxnote{why amsb? what is amsb?}, will have a chargino NLSP with an observable lifetime (\SI{0.2}{\nano\s} is a typical estimate\fxnote{why? how typical?}) that eventaully decays to the neutralino LSP and, commonly, a pion\fxnote{why?}. The nearly-degenerate masses have two important phenomenological roles: (1) increase the chargino lifetime, and (2) ensure low-momentum decay products. The low-momentum pion is then unlikely to be reconstructed, which results in a disappearing track~\cite{atlas_disappearing}.

@@

    Interpreting current limits in the context of naturalness requires some set of criteria that determine if a model is natural. No well-defined, universally-agreed-upon criteria exist, but certain themes recur in many discussions of natural model building\fxnote{better be able to confidently cite several sources}. The following discussion lays out a simplistic set of criteria for natural SUSY models and attempts to assess the status of natural SUSY in light of current experimental limits\fxnote{Did I just steal Chris's langauge?}.

@@

    \fxnote{cite cornering} 
    \fxnote{make sure these are logically and notationally consistent w/ earlier higgs corrections} 
    \noindent \begin{equation}
        \delta \mHiggs^2 = \mu^2
        \label{higgsino_correction}
    \end{equation}
    \noindent \begin{equation}
        \delta \mHiggs^2 \approx - \frac{3}{8\pi^2} y_t^2 m_{\tilde{t}}^2 \ln{\left(\frac{\Lambda}{Q}\right)}
        \label{stop_correction}
    \end{equation}
    \noindent \begin{equation}
        \delta \mHiggs^2 \approx - \frac{g_3^2 y_t^2}{4\pi^4} M_3^2 \ln{\left(\frac{\Lambda}{Q}\right)}^2
        \label{gluino_correction}
    \end{equation}

    Natural SUSY scenarios, then, are most likely to be constrained by limits on the stop, gluino, and Higgsino masses. The Higgsino, however, does not carry color charge and is far less likely\fxnote{be able to quantify} to be produced at the LHC, leaving stop and gluino mass limits as the most important contraints on natural SUSY models. As shown in Eqs.~\ref{stop_correction} and~\ref{gluino_correction}, the size of stop and gluino corrections also depends on $\Lambda$, the messenger scale of SUSY breaking.\fxnote{Make sure I understand this. Messenger scale != breaking scale?}\fxnote{Either explain more thoroughly or drop completely. What are the consequences of a lower messenger scale?}

    Two qualitative decisions must be made when analyzing the state of naturalness: (1) How best to define the degree of fine tuning in a particular model and (2) what degree of fine tuning is acceptable? A common, but not universal, definition for (1) is

    \fxnote{cite cornering} 
    \noindent \begin{equation}
        \Delta_i = \frac{\partial \ln{\mHiggs^2}}{\partial \ln{\alpha_i}} \quad ,
        \label{fine_tuning}
    \end{equation}

    \noindent where $\alpha_i$ is a particular parameter and $\Delta_i$ is the degree of fine tuning from variations in that parameter. The overall degree of fine tuning for the model can then be assigned as $\max{\left(\Delta_i\right)}$. Determining the acceptable degree of fine tuning is even more open ended. Before null results and serious constraints started rolling in, it was common\fxnote{cite something} to refer to the $\Delta \lesssim 10$ regions of parameter space as natural. As models are constrained, physicists seem willing to accept greater degrees of fine tuning. Instead of asking, ``what degree of fine tuning should be considered natural?", Feng\fxnote{cite} proposes physicists ask themselves, ``If SUSY were discovered with such-and-such degree of fine tuning, would I consider the naturalness problem solved and move on to different research questions?" Anecdotally, Feng reports that acceptable levels of fine tuning tend to increase when the question is posed in this way.\fxnote{break into two paragraphs?}

    So, how is natural SUSY fairing? Fig.~\ref{atlas_susy_searches} summarizes limits on many SUSY models and topologies from ATLAS searches. CMS limits are generally similar\fxnote{make sure this is true}, but a recent plot was not available.\fxnote{give general summary}

@@

    It's often helpful to look at limits in the context of simplified models. Fig.~\ref{cornering_limits} does just this for the cases of Effective SUSY, RPV SUSY, Hidden Valley SUSY, and a model that combines RPV with Effective SUSY.

@@

    Despite strong contraints, SUSY remains one of the least-tuned options for BSM naturalness\fxnote{Make sure I can explain this and make sure it isn't plagarized from Cornering}.

@@

\clearpage
\pagebreak
\singlespacing
\bibliography{draft}{}
\bibliographystyle{unsrt}
